---
title: "syntenic plots (code adapted from https://github.com/marta-fb/syntenyPlotteR)"
output:
  html_document:
    df_print: paged
    
author: Hernan Lorenzi
---

### Load required libraries
```{r}
library(tidyverse)
library(ggplot2)
library(colortools)
library(cowplot)
```
**Example of input files:**

1- syntenic_annotation.txt:

NC_027961.1	1	15264	scf_41	25392	40656	-1	99.85

NC_071294.1	346	5741097	scf_16	6910965	12651725	-1	99.99

NC_071294.1	5741098	12620423	scf_16	31704	6911000	-1	99.99

NC_071294.1	12620125	12651313	scf_16	814	32037	-1	99.76

2- new_genome_size:

NC_071294.1	12651329

NC_071295.1	12514254

NC_071296.1	12058152

3- old_genome_size:

scf_1	11091506

scf_2	10267549

scf_3	12481282

### Define input files and names
```{r}
# Init variables
infile <- "syntenic_annotation.txt"
output <- "syntenic_annotation_plot"
refSizes <- "new_genome_size"
tarSizes <- "old_genome_size"
refName <- "P. interpunctella\n(GCF_027563975.1)"
tarName <- "P. interpunctella\n(old)"
```


### Upload data
```{r}
xstart<-xend<-refchr<-tarchr<-x<-y<-group<-fill<-NULL
dataTMP<- tibble(read.delim(infile, header=FALSE))
data<-dataTMP[,c(4,5,6,1,2,3,7,8)]
ref_sizes <-tibble(read.delim(refSizes, header=FALSE)) #to be consistent with naming in EH
tar_sizes <-tibble(read.delim(tarSizes, header=FALSE))
colnames(data) = c("tarchr", "tarstart", "tarend", "refchr", "refstart", "refend", "dir", "notUsed")
colnames(ref_sizes) = c("refchr", "size")
colnames(tar_sizes) = c("tarchr", "size")

```

### Sort input data 
```{r}
# Sort ref scf by size
ref_sizes <- ref_sizes[order(ref_sizes$size, decreasing = TRUE),]

# Sort data df by ref_sizes positions
data_sorted <- left_join(data, ref_sizes, by = "refchr")
data_sorted <- data_sorted[order(data_sorted$size, decreasing = TRUE),]
data <- data_sorted

# Create vector with target scf ids
tar_scf_ids <- unique(data_sorted$tarchr)
missing_tar <- as.vector(tar_sizes$tarchr[! tar_sizes$tarchr %in% tar_scf_ids])
tar_scf_ids  <- c(tar_scf_ids,missing_tar)
tar_sizes <- left_join(data.frame(tarchr = tar_scf_ids), tar_sizes, by = "tarchr")

```

### Add gap in between reference chromosomes and convert to "linear" genome
```{r}

for (i in c(1:nrow(ref_sizes))){
  #print(i)
  if (i == 1){
    total_start = 1
    total_end = ref_sizes[i, "size"]
  } else {
    total_start = total_end + 6000000
    total_end = total_start + ref_sizes[i, "size"]
  }
  ref_sizes[i,"xstart"] = total_start
  ref_sizes[i, "xend"] = total_end
}
```

### Add gap in between target chromosomes
```{r}
for (i in c(1:nrow(tar_sizes))){
  #print(i)
  if (i == 1){
    total_start = 1
    total_end = tar_sizes[i, "size"]
  } else {
    total_start = total_end + 6000000
    total_end = total_start + tar_sizes[i, "size"]
  }
  tar_sizes[i,"xstart"] = total_start
  tar_sizes[i, "xend"] = total_end
}
```


### Convert coordinates to linear genome and creates synteny polygon coordinates
```{r}
#This converts coordinates to linear genome and creates synteny polygon coordinates
synteny = data.frame()
for (i in c(1:nrow(data))){
  tar_chr = data[[i,"tarchr"]] # [[]] is required for tibble, 
                              # otherwise returns a tibble and not a scalar
  ref_chr = data[[i,"refchr"]]
  dir = data[[i, "dir"]]
  
  tar_add = tar_sizes[as.character(tar_sizes$tarchr)==as.character(tar_chr),]$xstart
  ref_add = ref_sizes[as.character(ref_sizes$refchr)==as.character(ref_chr),]$xstart
  tar_y = 0.1
  ref_y = 2
  tar_xstart = data[[i,"tarstart"]] + tar_add
  tar_xend = data[[i,"tarend"]] + tar_add
  ref_xstart = data[[i,"refstart"]] + ref_add
  ref_xend = data[[i,"refend"]] + ref_add
  
  inverted = grepl("-", dir, fixed = TRUE)
  if(inverted == TRUE){
    df = data.frame(x = c(tar_xstart, tar_xend, ref_xstart, ref_xend), 
                    y = c(tar_y, tar_y, ref_y, ref_y),
                    fill = ref_chr,
                    group = paste0("s",i)
                    )
  } else {
    df = data.frame(x = c(tar_xstart, ref_xstart, ref_xend,  tar_xend), y = c(tar_y, ref_y, ref_y, tar_y),
                    fill = ref_chr, 
                    group = paste0("s",i))
  }
  synteny = rbind(synteny,df)
}
```

```{r}
#making sure chr columns are factors
tar_sizes$tarchr<-as.factor(tar_sizes$tarchr)
ref_sizes$refchr<-as.factor(ref_sizes$refchr)
synteny$fill<-as.factor(synteny$fill)
  
```

### Make plots

#### Set pallete
```{r}
my_pallete <- list()
my_pallete <- wheel("darkblue", num = nrow(ref_sizes))
names(my_pallete) <- ref_sizes$refchr

my_pallete_synt <- list()
my_pallete_synt <- sample(wheel("darkblue", num = length(unique(synteny$group))))
names(my_pallete_synt) <- (unique(synteny$group))

```

#### Generate synteny plot coloring by reference chromosome
```{r}
#This prints plot


pairwise.p1 <- ggplot(linewidth = 0.2, font = 10, data = data) +
    geom_rect(data=ref_sizes, 
              mapping=aes(xmin=xstart, 
                          xmax=xend, 
                          ymin=2, 
                          ymax=2.10, 
                          fill=refchr),
              color="black", alpha = 0.75, size = 0.3 
              ) +
    geom_text(data=ref_sizes,aes(x=(xstart+xend)/2,y=2.3,label=refchr),size=3,angle=45) +
    geom_text(mapping=aes(x=-5,y=2, label=refName),size=3,hjust = 1.5) +
    geom_rect(data=tar_sizes, mapping=aes(xmin=xstart, xmax=xend, ymin=0, ymax=0.10),fill="grey85",
                       color="black", alpha = 0.75, size = 0.2 ) +
    geom_text(data=tar_sizes,aes(x=(xstart+xend)/2,y=-0.2,label=tarchr),size=3,angle=45) +
    geom_text(mapping=aes(x=-5,y=0.2, label=tarName),size=3,hjust = 1.5) +
    geom_polygon(data = synteny, alpha = .5, aes(x = x, y = y, group = group, fill = fill)) +
    scale_fill_manual(values = my_pallete) +
    theme(panel.background = element_blank(),
                   strip.background = element_blank(),
                   axis.title.y = element_blank(),
                   axis.title.x = element_blank(),
                   axis.text.x = element_blank(),
                   axis.text.y = element_blank(),
                   axis.ticks.x=element_blank(),
                   axis.ticks.y=element_blank(),
                   legend.position="none")

ggsave2(filename = "synteny_plot.pdf", plot = pairwise.p1, width = 25, height = 5)
print(pairwise.p1)
  

```

#### Generate synteny plot coloring by syntenic block
```{r}
#This prints plot


pairwise.p2 <- ggplot(linewidth = 0.2, font = 10, data = data) +
    geom_rect(data=ref_sizes, 
              mapping=aes(xmin=xstart, 
                          xmax=xend, 
                          ymin=2, 
                          ymax=2.10, 
                          fill=refchr),
              color="black", alpha = 0.75, size = 0.3 
              ) +
    geom_text(data=ref_sizes,aes(x=(xstart+xend)/2,y=2.3,label=refchr),size=3,angle=45) +
    geom_text(mapping=aes(x=-5,y=2, label=refName),size=3,hjust = 1.5) +
    geom_rect(data=tar_sizes, mapping=aes(xmin=xstart, xmax=xend, ymin=0, ymax=0.10),fill="grey85",
                       color="black", alpha = 0.75, size = 0.2 ) +
    geom_text(data=tar_sizes,aes(x=(xstart+xend)/2,y=-0.2,label=tarchr),size=3,angle=45) +
    geom_text(mapping=aes(x=-5,y=0.2, label=tarName),size=3,hjust = 1.5) +
    geom_polygon(data = synteny, alpha = .5, aes(x = x, y = y, group = group, fill = group)) +
    scale_fill_manual(values = my_pallete_synt) +
    theme(panel.background = element_blank(),
                   strip.background = element_blank(),
                   axis.title.y = element_blank(),
                   axis.title.x = element_blank(),
                   axis.text.x = element_blank(),
                   axis.text.y = element_blank(),
                   axis.ticks.x=element_blank(),
                   axis.ticks.y=element_blank(),
                   legend.position="none")

ggsave2(filename = "synteny_plot_color_synt_block.pdf", plot = pairwise.p2, width = 25, height = 5)
print(pairwise.p2)
  

```






