---
title: "syntenic plots (code adapted from https://github.com/marta-fb/syntenyPlotteR)"
output: html
author: Hernan Lorenzi
---

### Load required libraries
```{r}
library(tidyverse)
library(ggplot2)
library(colortools)
library(cowplot)
```
**Example of input files:**

1- syntenic_annotation.txt:

NC_027961.1	1	15264	scf_41	25392	40656	-1	99.85

NC_071294.1	346	5741097	scf_16	6910965	12651725	-1	99.99

NC_071294.1	5741098	12620423	scf_16	31704	6911000	-1	99.99

NC_071294.1	12620125	12651313	scf_16	814	32037	-1	99.76

2- new_genome_size:

NC_071294.1	12651329

NC_071295.1	12514254

NC_071296.1	12058152

3- old_genome_size:

scf_1	11091506

scf_2	10267549

scf_3	12481282

### Define input files and names
```{r}
# Init variables
infile <- "Pinter_vs_Bmori_promer.matchList.filtered.synteny_chr.txt"
output <- "syntenic_annotation_Bm_Pi_plot"
refSizes <- "karyotype.txt"
tarSizes <- "new_genome_size.txt"
refName <- "B.mori"
tarName <- "P.interpunctella (new)"
chr2acc_file <- "chr2acc_Bmori.csv"
```




### Upload data
```{r}
xstart<-xend<-refchr<-tarchr<-x<-y<-group<-fill<-NULL
dataTMP<- tibble(read.delim(file =infile, header=TRUE))
data<-dataTMP[,c(4,5,6,1,2,3)]
data <- data %>% mutate( dir = ifelse(((as.numeric(End_2) - as.numeric(Start_2)) * (as.numeric(End_1) - as.numeric(Start_1)) ) > 0, 1, -1))

ref_sizes <-tibble(read.delim(refSizes, header=TRUE)) #to be consistent with naming in EH
tar_sizes <-tibble(read.delim(tarSizes, header=TRUE))
colnames(data) = c("tarchr", "tarstart", "tarend", "refchr", "refstart", "refend", "dir")
colnames(ref_sizes) = c("refchr", "size")
colnames(tar_sizes) = c("tarchr", "size")

# chr2acc is required if ref_sizes, tar_sizes and data files use accession numbers and you want to display chromosome IDs instead.
chr2acc<- tibble(read.delim(file = chr2acc_file, header=TRUE, sep = ","))

```

# Filter out small chromosomes
```{r}
# Select chromosomes to keep
cutoff = 500000
ref_long <- filter(ref_sizes, size > cutoff)$refchr
tar_long <- filter(tar_sizes, size > cutoff)$tarchr

# remove small chromosomes from syntenic blocks and tar/ref_size files
data <- filter(data, tarchr %in% tar_long & refchr %in% ref_long)
ref_sizes <- filter(ref_sizes, refchr %in% ref_long)
tar_sizes <- filter(tar_sizes, tarchr %in% tar_long)
```

### Select specifc target chromosomes for plots (leave as is for all genome)
```{r}
target_chr_id = 'all'
if(target_chr_id != 'all'){
  target_chrom <- c(target_chr_id)
  data <- filter(data,tarchr %in% target_chrom)
  ref_sizes <-  filter(ref_sizes, refchr %in% data$refchr)
  tar_sizes <- filter(tar_sizes, tarchr %in% target_chrom)
}
```

### Sort input data 
```{r}
# Sort ref scf by size
ref_sizes <- ref_sizes[order(ref_sizes$size, decreasing = TRUE),]

# Sort data df by ref_sizes positions
data_sorted <- left_join(data, ref_sizes, by = "refchr")
data_sorted <- data_sorted[order(data_sorted$size, decreasing = TRUE),]
data <- data_sorted

# Create vector with target scf ids

data_max <- data %>% mutate(hit_len = refend - refstart) %>% 
                     group_by(refchr) %>%
                     filter(hit_len == max(hit_len))
tar_scf_ids <- unique(data_max$tarchr)
#tar_scf_ids <- unique(data_sorted$tarchr)
missing_tar <- as.vector(tar_sizes$tarchr[! tar_sizes$tarchr %in% tar_scf_ids])
tar_scf_ids  <- c(tar_scf_ids,missing_tar)
tar_sizes <- left_join(data.frame(tarchr = tar_scf_ids), tar_sizes, by = "tarchr")

```

## Sort tar_sizes based on  data$refchr order and syntenic blocks##
```{r TEMP}
x <- filter(data %>% group_by(refchr,tarchr) %>% count(), n > 5)

# Create numerical index list for data$refchr to match chr order
my_ref_chr_order <- 1:length(unique(data$refchr))
names(my_ref_chr_order) <- unique(data$refchr)
                    
# Sort Bm chr in x based on my_Bm_chr_order
x$order_idx <- my_ref_chr_order[x$refchr]
x <- x[order(x$order_idx, decreasing = F),]

# Sort tar_sizes based on x order
my_tar_chr_order <- 1:nrow(x)
names(my_tar_chr_order) <- x$tarchr
tar_sizes$order_idx <- my_tar_chr_order[tar_sizes$tarchr]
tar_sizes <- tar_sizes[order(tar_sizes$order_idx, decreasing = F),]
```

### Add gap in between reference chromosomes and convert to "linear" genome
```{r}

for (i in c(1:nrow(ref_sizes))){
  #print(i)
  if (i == 1){
    total_start = 1
    total_end = ref_sizes[i, "size"]
  } else {
    total_start = total_end + 6000000
    total_end = total_start + ref_sizes[i, "size"]
  }
  ref_sizes[i,"xstart"] = total_start
  ref_sizes[i, "xend"] = total_end
}
```

### Add gap in between target chromosomes
```{r}
for (i in c(1:nrow(tar_sizes))){
  #print(i)
  if (i == 1){
    total_start = 1
    total_end = tar_sizes[i, "size"]
  } else {
    total_start = total_end + 6000000
    total_end = total_start + tar_sizes[i, "size"]
  }
  tar_sizes[i,"xstart"] = total_start
  tar_sizes[i, "xend"] = total_end
}
```

## Calculate left padding to center ref and tar chromosomes
```{r}
plot_align_mode <- c('centered','spanned')[2] # set to 1 or 2
tar_pad <- ref_pad <- 0
ref_len <- tail(ref_sizes$xend,1)
tar_len <- tail(tar_sizes$xend,1)


if (ref_len > tar_len){
  if (plot_align_mode == 'centered'){
      tar_pad <- round((ref_len - tar_len) / 2, digits = 0)
      tar_sizes$xstart <- tar_sizes$xstart + tar_pad
      tar_sizes$xend <- tar_sizes$xend + tar_pad
    } else {
      n_chrom <- nrow(tar_sizes)
      tar_pad <- round((ref_len - tar_len) / (n_chrom - 1), digits = 0)
      p = 0:(n_chrom - 1)
      tar_pad_vector <- tar_pad * p
      tar_sizes$xstart <- tar_sizes$xstart + tar_pad_vector
      tar_sizes$xend <- tar_sizes$xend + tar_pad_vector
    }
} else {
  if (plot_align_mode == 'centered'){
      ref_pad <- round((tar_len - ref_len) / 2, digits = 0)
      ref_sizes$xstart <- ref_sizes$xstart + ref_pad
      ref_sizes$xend <- ref_sizes$xend + ref_pad
    } else {
      n_chrom <- nrow(ref_sizes)
      ref_pad <- round((tar_len - ref_len) / (n_chrom - 1), digits = 0)
      p = 0:(n_chrom - 1)
      ref_pad_vector <- ref_pad * p
      ref_sizes$xstart <- ref_sizes$xstart + ref_pad_vector
      ref_sizes$xend <- ref_sizes$xend + ref_pad_vector
  }
}
```

### Calculate target chromosome orientation
```{r}
# Reverse syntenic blocks if hits are mostly on reverse strand
chr_strand_target = 'forward'
if (!is.na(table(data$dir == -1)['TRUE']) & ( (table(data$dir == -1)['TRUE'] - length(data$dir)) >= 0) ){
    chr_strand_target = 'reverse'
  
}

#> data$tarstart <- tar_sizes$size - data$tarstart
#> data$tarend <- tar_sizes$size - data$tarend

```

### Convert coordinates to linear genome and creates synteny polygon coordinates
```{r}
#This converts coordinates to linear genome and creates synteny polygon coordinates
synteny = data.frame()
for (i in c(1:nrow(data))){
  tar_chr = data[[i,"tarchr"]] # [[]] is required for tibble, 
                              # otherwise returns a tibble and not a scalar
  ref_chr = data[[i,"refchr"]]
  dir = data[[i, "dir"]]
  
  # Adjust position when there are many chromosomes
  tar_add = tar_sizes[as.character(tar_sizes$tarchr)==as.character(tar_chr),]$xstart
  ref_add = ref_sizes[as.character(ref_sizes$refchr)==as.character(ref_chr),]$xstart
  tar_y = 0.05
  ref_y = 2.05
  tar_xstart = data[[i,"tarstart"]] + tar_add
  tar_xend = data[[i,"tarend"]] + tar_add
  ref_xstart = data[[i,"refstart"]] + ref_add
  ref_xend = data[[i,"refend"]] + ref_add
  
  df = data.frame(x = c(tar_xstart, ref_xstart, ref_xend,  tar_xend), y = c(tar_y, ref_y, ref_y, tar_y),fill = ref_chr, group = paste0("s",i))
  
  synteny = rbind(synteny, df)
  
  # Substract tar_chr size to tar_choords for inverted syntenic block
  inverted = grepl("-", dir, fixed = TRUE)
  if(inverted == TRUE & chr_strand_target == "reverse"){
    synteny[synteny$y == 0.05 & synteny$group == paste0("s",i), ]$x <- tar_sizes$size + 1 - synteny[synteny$y == 0.05 & synteny$group == paste0("s",i), ]$x
  }  
}
```

```{r}
#making sure chr columns are factors
tar_sizes$tarchr<-as.factor(tar_sizes$tarchr)
ref_sizes$refchr<-as.factor(ref_sizes$refchr)
synteny$fill<-as.factor(synteny$fill)
  
```

### Make plots

#### Set pallete
```{r}
my_pallete <- list()
my_pallete <- wheel("darkblue", num = nrow(ref_sizes))
names(my_pallete) <- ref_sizes$refchr

my_pallete_synt <- list()
my_pallete_synt <- sample(wheel("darkblue", num = length(unique(synteny$group))))
names(my_pallete_synt) <- (unique(synteny$group))

```

#### Generate synteny plot coloring by reference chromosome
```{r}

# Function to create scale
get_chr_scale <- function(chr_size, position, strand = 'forward'){
  if(position == 'top'){
    y_min <- 2.1
    y_med <- 2.12
    y_max <- 2.14
  }else if(position == 'bottom'){
    y_min <- 0
    y_med <- -0.02
    y_max <- -0.04
  }else{
    stop()
  }
  divisions <- 2.5e5
  numb_div <- round(chr_size/divisions, digits = 0)
  x_coord <- c()
  if(strand=='forward'){
    counter <- 0
    chr_pos <- c()
    count_pos <- 0
    while (counter <=  chr_size){
      x_coord <- c(x_coord, counter)
      counter <- counter + divisions
      chr_pos <- c(chr_pos, count_pos)
      count_pos <- count_pos + divisions
    }
  } else if(strand=='reverse'){
    counter <- chr_size
    chr_pos <- c()
    count_pos <- 0
    while (counter >= 0){
      x_coord <- c(x_coord, counter)
      counter <- counter - divisions
      chr_pos <- c(chr_pos, count_pos)
      count_pos <- count_pos + divisions
    }   
  } 
  chrom_pos <-  tibble(x_coord=x_coord, 
                       chr_pos=chr_pos,
                       y_start=y_min) 
  chrom_pos <- chrom_pos %>% mutate(y_end = ifelse(chr_pos / 1e6 == as.integer(chrom_pos$chr_pos / 1e6),y_max, y_med))
  chrom_pos <- chrom_pos %>% mutate(label=ifelse(chr_pos / 1e6 == as.integer(chrom_pos$chr_pos / 1e6),as.character(as.integer(chrom_pos$chr_pos / 1e6)), ""))
  return(chrom_pos)
}                    
```

```{r}
#This prints plot

if(target_chr_id != 'all'){
  my_chr_text_angle <- ifelse(length(unique(ref_sizes$refchr)) == 1, 0,45)
  
  pairwise.p1 <- ggplot(linewidth = 0.2, font = 10, data = data) + 
      geom_polygon(data = synteny, alpha = .5, 
                   aes(x = x, y = y, group = group, fill = fill)) +
      ggchicklet:::geom_rrect(data=ref_sizes, 
                mapping=aes(xmin=xstart, 
                            xmax=xend, 
                            ymin=2, 
                            ymax=2.10, 
                            fill=refchr),
                r = unit(0.5, 'npc'),
                color="black", 
                alpha = 1, 
                size = 0.3) 
  
  # Reference chromosome
  pairwise.p1 <- pairwise.p1 + 
      geom_segment(data = get_chr_scale(chr_size = ref_sizes$xend, position = 'top'), mapping = aes(x=x_coord, xend=x_coord, y=y_start, yend=y_end)) +
      geom_text(data=get_chr_scale(ref_sizes$xend, position = 'top', strand = 'forward'),
                aes(x=x_coord,y=y_end+0.02,label=label),size=3) +
      geom_text(data=ref_sizes,aes(x=(xstart+xend)/2,y=2.3,label=refchr),size=3,angle=my_chr_text_angle) +
      geom_text(mapping=aes(x=-5,y=2, label=refName),size=3,hjust = 1.5) 
  
  # Target chromosome
  pairwise.p1 <- pairwise.p1 +
      ggchicklet:::geom_rrect(data=tar_sizes, 
                mapping=aes(xmin=xstart, xmax=xend, ymin=0, ymax=0.10),
                r = unit(0.5, 'npc'),
                fill="grey85",
                color="black", 
                alpha = 1, 
                size = 0.2 ) +
      geom_segment(data = get_chr_scale(chr_size = tar_sizes$xend, # scale lines
                                        position = "bottom", 
                                        strand = chr_strand_target), 
                   mapping = aes(x=x_coord, xend=x_coord, y=y_start, yend=y_end)) +
      geom_text(data=get_chr_scale(tar_sizes$xend, # scale text
                                   position = 'bottom', 
                                   strand = chr_strand_target),
                mapping = aes(x=x_coord,y=y_end-0.02,label=label),size=3) +
      geom_text(data=tar_sizes, # target chrom label
                aes(x=(xstart+xend)/2,y=-0.2,label=tarchr),
                size=3,angle=my_chr_text_angle) +
      geom_text(mapping=aes(x=-5,y=0.2, label=tarName),size=3,hjust = 1.5)
  
  # Polygon
  # z$x[z$y == 0.1] <- tar_sizes$size + 1 - subset(z, y == 0.1)$x
  pairwise.p1 <- pairwise.p1 + scale_fill_manual(values = my_pallete) 
  
  # Theme
  pairwise.p1 <- pairwise.p1 +
      theme(panel.background = element_blank(),
                     strip.background = element_blank(),
                     axis.title.y = element_blank(),
                     axis.title.x = element_blank(),
                     axis.text.x = element_blank(),
                     axis.text.y = element_blank(),
                     axis.ticks.x=element_blank(),
                     axis.ticks.y=element_blank(),
                     legend.position="none")
      
  ggsave2(filename = paste0("synteny_plot_",target_chr_id,".pdf"), plot = pairwise.p1, width = 8, height = 5)
  print(pairwise.p1)
  
}
```

#### Generate synteny plot coloring by syntenic block
```{r}
#This prints plot

if(target_chr_id != 'all'){
  pairwise.p2 <- ggplot(linewidth = 0.2, font = 10, data = data) +
    geom_polygon(data = synteny, alpha = .5, aes(x = x, y = y, group = group, fill = group)) +
      ggchicklet:::geom_rrect(data=ref_sizes, 
                mapping=aes(xmin=xstart, 
                            xmax=xend, 
                            ymin=2, 
                            ymax=2.10, 
                            fill=refchr),
                r = unit(0.5, 'npc'),
                color="black", 
                alpha = 1, 
                size = 0.3 
                ) +
      geom_segment(data = get_chr_scale(ref_sizes$xend, position = 'top', strand = 'forward'), mapping = aes(x=x_coord, xend=x_coord, y=y_start, yend=y_end)) +
      geom_text(data=get_chr_scale(ref_sizes$xend, position = 'top', strand = 'forward'),
                aes(x=x_coord,y=y_end+0.02,label=label),size=3) +
      geom_text(data=ref_sizes,aes(x=(xstart+xend)/2,y=2.3,label=refchr),size=3,angle=my_chr_text_angle) +
      geom_text(mapping=aes(x=-5,y=2, label=refName),size=3,hjust = 1.5) +
      ggchicklet:::geom_rrect(data=tar_sizes, 
                              mapping=aes(xmin=xstart, xmax=xend, ymin=0, ymax=0.10),
                              r = unit(0.5, 'npc'),
                              fill="grey85",
                              color="black", 
                              alpha = 1, 
                              size = 0.2 ) +
      geom_segment(data = get_chr_scale(tar_sizes$xend, position = "bottom", strand = chr_strand_target), mapping = aes(x=x_coord, xend=x_coord, y=y_start, yend=y_end)) +
      geom_text(data=get_chr_scale(tar_sizes$xend, position = 'bottom', strand = chr_strand_target),
                aes(x=x_coord,y=y_end-0.02,label=label),size=3) +
      geom_text(data=tar_sizes,aes(x=(xstart+xend)/2,y=-0.2,label=tarchr),size=3,angle=my_chr_text_angle) +
      geom_text(mapping=aes(x=-5,y=0.2, label=tarName),size=3,hjust = 1.5) +
      scale_fill_manual(values = my_pallete_synt) +
      theme(panel.background = element_blank(),
                     strip.background = element_blank(),
                     axis.title.y = element_blank(),
                     axis.title.x = element_blank(),
                     axis.text.x = element_blank(),
                     axis.text.y = element_blank(),
                     axis.ticks.x=element_blank(),
                     axis.ticks.y=element_blank(),
                     legend.position="none")
  
  ggsave2(filename = paste0("synteny_plot_color_synt_block_",target_chr_id,".pdf"), plot = pairwise.p2, width = 8, height = 5)
  print(pairwise.p2)
  
}
```

#### Generate synteny plot coloring by reference chromosome
```{r}
#This prints plot
if(target_chr_id == 'all'){
  pairwise.p1 <- ggplot(linewidth = 0.2, font = 10, data = data) +
      geom_polygon(data = synteny, alpha = 0.8, aes(x = x, y = y, group = group, fill = fill)) +
      geom_rect(mapping = aes(xmin=-40, xmax=-3, ymin=-1.5, ymax=3.5),
              color="white", 
              alpha = 0, 
              size = 0.3 ) +
      ggchicklet:::geom_rrect(data=ref_sizes, 
                              mapping=aes(xmin=xstart, 
                                          xmax=xend, 
                                          ymin=2, 
                                          ymax=2.10, 
                                          fill=refchr),
                              r = unit(0.5, 'npc'),
                              color="black", 
                              alpha = 1, 
                              size = 0.3 ) + 
      geom_text(data=ref_sizes,aes(x=(xstart+xend)/2,y=2.15,label=refchr),
                size=3,
                angle=90, 
                hjust = 0) +
      #geom_text(mapping=aes(x=-5,y=2, label=refName),size=3,hjust = 1.5) +
      ggchicklet:::geom_rrect(data=tar_sizes, 
                              mapping=aes(xmin=xstart, xmax=xend, ymin=0, ymax=0.10),
                              r = unit(0.5, 'npc'),
                              fill="grey85",
                              color="black", 
                              alpha = 1, 
                              size = 0.2 ) +
      geom_text(data=tar_sizes,aes(x=(xstart+xend)/2,y=-0.1,label=tarchr),
                              size=3,
                              angle=90,
                              hjust = 1) +
      #geom_text(mapping=aes(x=-5,y=0.2, label=tarName),size=3,hjust = 1.5) +
      scale_fill_manual(values = my_pallete) +
      theme(panel.background = element_blank(),
                     strip.background = element_blank(),
                     axis.title.y = element_blank(),
                     axis.title.x = element_blank(),
                     axis.text.x = element_blank(),
                     axis.text.y = element_blank(),
                     axis.ticks.x=element_blank(),
                     axis.ticks.y=element_blank(),
                     legend.position="none")
 
            
  ggsave2(filename = "synteny_plot.pdf", plot = pairwise.p1, width = 25, height = 4)
  print(pairwise.p1)
}
```

#### Generate synteny plot coloring by syntenic block
```{r}
#This prints plot
if(target_chr_id == 'all'){
  pairwise.p2 <- ggplot(linewidth = 0.2, font = 10, data = data) +
      geom_polygon(data = synteny, alpha = .5, aes(x = x, y = y, group = group, fill = group)) +
      geom_rect(mapping = aes(xmin=-30, xmax=-3, ymin=-1.5, ymax=3.5),
              color="white", 
              alpha = 0, 
              size = 0.3 ) +
      ggchicklet:::geom_rrect(data=ref_sizes, 
                mapping=aes(xmin=xstart, 
                            xmax=xend, 
                            ymin=2, 
                            ymax=2.10, 
                            fill=refchr),
                r = unit(0.5, 'npc'),
                color="black", 
                alpha = 1, 
                size = 0.3 ) +
      geom_text(data=ref_sizes,aes(x=(xstart+xend)/2,y=2.17,label=refchr),
                size=3,
                angle=90, 
                hjust = 0) +
      #geom_text(mapping=aes(x=-5,y=2, label=refName),size=3,hjust = 1.5) +
      ggchicklet:::geom_rrect(data=tar_sizes, 
                              mapping=aes(xmin=xstart, xmax=xend, ymin=0, ymax=0.10),
                              r = unit(0.5, 'npc'),
                              fill="grey85",
                         color="black", 
                         alpha = 1, 
                         size = 0.2 ) +
      geom_text(data=tar_sizes,aes(x=(xstart+xend)/2,y=-0.3,label=tarchr),
                size=3,
                angle=90,
                hjust = 1) +
      #geom_text(mapping=aes(x=-5,y=0.2, label=tarName),size=3,hjust = 1.5) +
      scale_fill_manual(values = my_pallete_synt) +
      theme(panel.background = element_blank(),
                     strip.background = element_blank(),
                     axis.title.y = element_blank(),
                     axis.title.x = element_blank(),
                     axis.text.x = element_blank(),
                     axis.text.y = element_blank(),
                     axis.ticks.x=element_blank(),
                     axis.ticks.y=element_blank(),
                     legend.position="none")
  ggsave2(filename = "synteny_plot_color_synt_block.pdf", plot = pairwise.p2, width = 25, height = 4)
  print(pairwise.p2)
}
```





